Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2018-02-23 paper.tex
tables updated, experiments reorganized

\author{Vojt\v ech Von\' asek$^{1}$,
    Barbora Kozl\'\i kov\'a$^{2}$,
    Adam Jur\v{c}\'\i k$^{2}$,
    Ond\v{r}ej V\'{a}vra$^{3}$
%    Martin Saska$^{1}$
\thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
Technick\'a 2, 166 27 Prague, Czech Republic, {\tt vonasek@labe.felk.cvut.cz};
$^{2}$ Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic;
$^{3}$ Loschmidt Laboratories, Department of Experimental Biology RECETOX, Faculty of Science, Masaryk University
    Kamenice 5, 625 00 Brno, Czech Republic. 
The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
}}
%The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
%Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
%}%
%}
%}

@@

\begin{abstract}
Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
These active sites are accessible through one or more void paths, called tunnels.
Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere and 
the traversability of the tunnels is determined based only on the radial bottlenecks.
This can lead to highly imprecise results.
To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations for flexible ligands.
To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
%which serves us for better localization of the searched space.
The flexibility of the ligand is modeled using a fixed set of known conformations, which does not increase the number of dimensions of the configuration space.
The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
\end{abstract}

@@

\section{Proposed Method}
%To address the problems of the existing solutions for computing the non-spherical flexible ligand trajectories through protein tunnels, we propose a modification of the RRT planner. 
We propose a modification of RRT to compute trajectories for non-spherical flexible ligands through protein tunnels.
%Our approach takes into account the shape and flexibility of the ligand and localizes the search around given tunnels. configuration space to 
%using the information about tunnels detected by the Voronoi-based approach.
The input is formed by a set of tunnels computed using Voronoi diagrams and the task is to find trajectories for a ligand along these tunnels
leading from outside the protein to the active site.
The tunnels are detected in a sequence of time frames of MD simulations.
%capturing protein dynamic behavior. 
%The frames are computed using MD simulations.

@@

Molecular dynamics simulations of ligands passing through proteins show that the ligands can be highly flexible.
However, ligands cannot form arbitrary conformations. 
The ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
Therefore, the ligand flexibility can be modeled using only a predefined set of conformations.
By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
the dimension of the configuration space is not increased as in~\cite{cortes2010simulating}.
%So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.

@@

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (its geometric center), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$, and the scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
We assume that the ligand can change from conformation $l_1$ at configuration $q_1$ to a conformation $l_2$ at configuration $q_2$
if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is less than the resolution of the planner ($\varepsilon=0.05$~\AA).

@@

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used. %~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
This allows the tree to slightly detour from the tunnel and overcome its difficult (i.e., narrow) parts by finding an alternative trajectory nearby.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

The movements of the ligands inside a tunnel are blocked by many atoms forming the tunnel.
It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
should be allowed, which is controlled by the $\dt$ parameter.
This parameter should be set considering the existence of cavities or other tunnels around the tunnel being investigated.
In the case of alone tunnels (which is very common), we recommend to use double value of the tunnel average width.

@@

The proposed method was used to compute the ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (PDB ID 4E46) which consists of 4650 atoms.
The active site was defined using residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid).
MD of the protein was computed using AMBER~12 (details about the used MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
In each frame, tunnels for the spherical probe 0.9~\AA\ were detected using CAVER 3.0~\cite{caver3}. 
In most of the frames, two tunnels were detected.
The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.
The tunnel traversability was analyzed in 100 consecutive frames of MD simulation in part where the protein was stabilized.

@@

\subsection{Traversability of One Atom}

The performance of \RA\ and \RB\ planners was first measured by finding trajectories for only one atom of VDW radii $1.7$~\AA.
The \RC\ variant was not tested here as it behaves same as \RA\ in the case of one-atom ligand.
The computed traversability rates are shown in Tab.~\ref{tab::one}.
The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
The row 'Planning' shows the traversable rate computed by the planners.
This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
increase the planning time.
For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.

\begin{table}[t]
\centering
\caption{\label{tab::one}
    \small
    Traversability rate [\%] of one atom.
}
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphProbe.tex
}
\end{table}

%\begin{figure}
%\centering
%\begin{tabular}{cc}
%\includegraphics[width=0.3\textwidth]{fig/t05proteintunnels} &
%\includegraphics[width=0.45\textwidth]{fig/ligAll}  \\
%(a) & (b)                        
%\end{tabular}                       
%\caption{\label{fig::tunnel}
%    (a) The protein 1CQW visualized using the cartoon representation (gray) with 
%        three tunnels (red, green, blue) detected by CAVER 3.0.
%        The first tunnel is depicted red.
%%        its highest ranked tunnel (green) which was detected using CAVER 3.0.
%    (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
%}
%\end{figure}
%

\subsection{Influence of the Conformation Selection}

The aim of this experiment was to verify the procedure to select feasible ligand conformations from an available pool of conformations.
Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075) 
Two strategies to select a subset from the $\L$ conformations were tested: 
the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller conformations (denoted `P').
The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
to the selection of packed conformations.
The strategies were tested using the ligand DCP (Dichloropropanol), for which $1300$ conformations were created, and for the ligand
m040 (1,5-Dichloropentane) with $7560$ conformations.

@@

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
    The number after~`$/$' denotes the number of atoms.
    The protein has 4650 atoms. 
    The rate is shown in [\%].
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphSelNew.tex
}
\end{table}

\subsection{Traversability of Ligands}

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
    The protein has 4650 atoms.
    The rate is shown in [\%].
}
\small
\renewcommand{\tabcolsep}{2pt}
{
\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphNew.tex
}
\end{table}

@@

\begin{table}[t]
\caption{\label{fig::m040c}
\small
Examples of elongated/packed conformation for m040 (1,5-Dichloropentane).
The table shows the average number of conformations (out of 50) used in trajectories reaching the active site.
}
\centering
{\footnotesize
\def\arraystretch{0.9}
\begin{tabular}{cc}
\rotatebox{0}{\hskip 5pt Long} \\ \includegraphics[width=0.1\textwidth]{fig/m040-conf1} \\
\rotatebox{0}{Packed} \\ \includegraphics[width=0.078\textwidth]{fig/m040-conf2}
\end{tabular}
}
{\scriptsize
\def\arraystretch{0.9}
\input usedc.tex
}
\end{table}

@@

%originally last two paragraphs in the experiment section, for resubmition, they are moved here to discussion
The proposed \RA\ planner provides higher traversability rate than RRT-Retraction (\RB).
The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
are shown in Fig.~\ref{fig::comparison}.
The runtimes of \RB\ are higher than runtimes of \RA and the std. deviations as well.
The variant \RC\ is faster than \RA\ due to simple evaluation of the metric in the expansion step.
%The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
%to the active site is more than 2~\AA\ (for example, in the frame no. 60).
The utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
%Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).

@@

We presented a novel modification of the RRT planner to compute trajectories of ligands inside selected protein tunnels.
In our solution, we take into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
The planner generates the random samples only along a tunnel computed by the Voronoi diagram, and it attempts to expand the configuration
tree using all ligand conformations. 
The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if necessary, i.e., when traversing the narrow parts of the tunnel.
To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.

>>>>

\author{Vojt\v ech Von\' asek$^{1}$,
    Barbora Kozl\'\i kov\'a$^{2}$,
    Adam Jur\v{c}\'\i k$^{2}$,
    Ond\v{r}ej V\'{a}vra$^{3}$
    Martin Saska$^{1}$
\thanks{$^{1}$ Faculty of Electrical Engineering,  Czech Technical University in Prague, 
Technick\'a 2, 166 27 Prague, Czech Republic, {\tt vonasek@labe.felk.cvut.cz};
$^{2}$ Faculty of Informatics, Masaryk University, Botanick\'a 68a, 602~00 Brno, Czech Republic;
$^{3}$ Loschmidt Laboratories, Department of Experimental Biology RECETOX, Faculty of Science, Masaryk University
    Kamenice 5, 625 00 Brno, Czech Republic. 
The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
}}
%The presented work has been supported by the Czech Science Foundation (GA{\v C}R) under research project No. 17-07690S.
%Access to computing and storage facilities owned by parties and projects contributing to the National Grid Infrastructure MetaCentrum, provided under the programme "Projects of Large Infrastructure for Research, Development, and Innovations" (LM2010005), is greatly appreciated.
%}%
%}
%}

@@

\begin{abstract}
Chemical interactions between proteins and other molecules (ligands) take place in specific locations, called active sites, that are often deeply buried inside the protein structure.
These active sites are accessible through one or more void paths, called tunnels.
Nowadays, tunnels are mostly computed using Voronoi diagrams which approximate the ligand by a bounding sphere and 
the traversability of the tunnels is determined based only on the radial bottlenecks.
This can lead to highly imprecise results.
To address this problem, we define the computation of the transportation path of non-spherical ligands including their conformation changes (i.e., when the atoms of the ligand move relatively to each other) as a motion planning task in the high-dimensional configuration space.
We propose a modification of the Rapidly Exploring Random Tree planner for the purpose of trajectory computations for flexible ligands.
To cope with the narrow passage problem, the random samples are generated around a tunnel detected by a Voronoi-based approach.
%which serves us for better localization of the searched space.
The flexibility of the ligand is modeled using a fixed set of known conformations. 
%which does not increase the number of dimensions of the configuration space.
The proposed method has been evaluated in a virtual screening scenario to estimate the traversability of tunnels for several ligands.
\end{abstract}

@@

\section{Proposed Method}
We propose a modification of RRT to compute trajectories for non-spherical flexible ligands through protein tunnels.
The input of the proposed method is a set of tunnels computed using Voronoi diagrams, and the task is to find trajectories for a ligand along these tunnels leading from outside the protein to the active site.
The tunnels are detected on selected protein conformations resulting e.g. from MD simulations or from tool providing conformation sampling
~\cite{lei2004sampling,wells2005constrained,seeliger2007geometry}.
%The tunnels are detected in a sequence of time frames of MD simulations.
%capturing protein dynamic behavior. 
%The frames are computed using MD simulations.

@@

Molecular dynamics simulations of ligands passing through proteins show that the ligands can be highly flexible.
However, ligands cannot form arbitrary conformations. 
The ligand can contain several torsion and dihedral angles between its atoms which determine possible ligand conformations.
Therefore, the ligand flexibility can be modeled using only a predefined set of conformations.
By using the set of predefined conformations, it is not necessary to model the additional degrees of freedom, so 
the dimension of the configuration space is not increased as in~\cite{cortes2010simulating}.
%So the number of dimensions of the configuration space does not depend on the DOF of the ligand as in~\cite{cortes2010simulating}.
%The libraries containing possible ligand conformations are used also in different types of calculations, e.g., in~\cite{kellogg}.

@@

A configuration of the ligand $q=(x,y,z,r_x,r_y,r_z,l,s)$ is described
by the 3D position $(x,y,z)$ of the reference point of the ligand (e.g. the geometric center or a selected atom), the rotation around $x$, $y$, and $z$ axes,
the index of the conformation $l\in \L$, and the scale $s \in \S$.
All possible configurations form the configuration space $\C$ and the collision-free configurations
form the subspace $\CF \subseteq \C$.
A configuration $q$ is collision-free if none of the ligand atoms scaled by $s$ and placed at the
position defined by $q$ collides with the protein atoms.
We assume that the ligand can change from conformation $l_1$ at configuration $q_1$ to a conformation $l_2$ at configuration $q_2$
if both $q_1$ and $q_2$ are collision-free and if the distance from $q_1$ to $q_2$ is less than the resolution of the planner ($\varepsilon=0.05$~\AA).

@@

The aim of the first modification is to sample the configuration space along the tunnel being analyzed.
To guide the growth of the tree using the tunnel, a moving virtual goal is used. %~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a sphere of the tunnel.
At the beginning of the algorithm, $v=1$.
The random samples $\qrand$ are generated around the sphere $c_v \in T$ with the probability $\gb$, and from the whole $\C$ otherwise.
After the tree reaches the sphere $c_v$, i.e., the distance of the tree to $c_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to the successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are not traversable or reachable by the ligand.
The tree can slightly detour from the tunnel and overcome difficult parts (e.g., the tunnel bottleneck) by finding an alternative trajectory nearby.
%It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
%should be allowed, which is controlled by the $\dt$ parameter.
%parameter should be set considering the existence of cavities or other tunnels around the tunnel being investigated.
The parameter $\dt$ determines how far can the tree detour from the tunnel centerline.
In the case of alone tunnels (which is very common), we recommend to use double value of the tunnel average width.
The algorithm terminates after a predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

%The movements of the ligands inside a tunnel are blocked by many atoms forming the tunnel.

@@

The proposed algorithm was verified in two scenarios: on a sequence frames of MD simulation and on a set of protein conformations provided by the tCONCOORD tool~\cite{seeliger2007geometry}.

\subsection{Analysis of MD simulation}

\subsubsection{Algorithm and data setup}

The task is to compute the ligand traversability towards the active site using two main tunnels in the Haloalkane dehalogenase protein (DhaAwt, PDB ID 4E46) which consists of 4650 atoms.
MD of the protein was computed using AMBER~12 (details about the used MD simulations of DhaAwt/4E46 are described in~\cite{marques2017catalytic}).
The tunnels were detected in 100 consecutive frames of MD simulation in part where the protein was stabilized.
The tunnels for probe 0.9~\AA\ towards the active site defined by residues 38 (Asparagine), 102 (Histidine), and 103 (Aspartic Acid) were
computed using the CAVER 3.0~\cite{caver3} tool.
In most of the frames, two tunnels were detected.
The average bottleneck sizes of these tunnels are $1.48$~\AA\ and $1.2$~\AA\ for the main and the side tunnel, respectively.

Models  of the ligands in Mol2 formats were converted to param files~\cite{meiler2006rosettaligand}, containing the topology, rotatable bonds, atom types, and partial charges. 
The pool of conformations $\L$ was created by the rotation of all rotatable bonds with 30$^\circ$ step using Rosetta3.
The energy cut-off for individual conformations was set to 0.075. %(-rot_ensemble_ecutoff 0.075)

@@

%\subsection{Traversability of One Atom}
%
%The performance of \RA\ and \RB\ planners was first measured by finding trajectories for only one atom of VDW radii $1.7$~\AA.
%The \RC\ variant was not tested here as it behaves same as \RA\ in the case of one-atom ligand.
%The computed traversability rates are shown in Tab.~\ref{tab::one}.
%The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
%The row 'Planning' shows the traversable rate computed by the planners.
%This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
%The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
%Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
%increase the planning time.
%For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.
%
%\begin{table}[t]
%\centering
%\caption{\label{tab::one}
%    \small
%    Traversability rate [\%] of one atom.
%}
%\renewcommand{\tabcolsep}{4.3pt}
%{\scriptsize
%\renewcommand{\arraystretch}{0.7}
%\input graphProbe.tex
%}
%\end{table}
%
%
%
%%\begin{figure}
%%\centering
%%\begin{tabular}{cc}
%%\includegraphics[width=0.3\textwidth]{fig/t05proteintunnels} &
%%\includegraphics[width=0.45\textwidth]{fig/ligAll}  \\
%%(a) & (b)                        
%%\end{tabular}                       
%%\caption{\label{fig::tunnel}
%%    (a) The protein 1CQW visualized using the cartoon representation (gray) with 
%%        three tunnels (red, green, blue) detected by CAVER 3.0.
%%        The first tunnel is depicted red.
%%%        its highest ranked tunnel (green) which was detected using CAVER 3.0.
%%    (b) Examples of three conformations of $\LA$ (top) and $\LB$ (bottom)
%%}
%%\end{figure}
%%
%

\subsubsection{Influence of the Conformation Selection}

The number of generated ligand conformations $|\L|$ depends on the number of rotatable bonds of the ligands.
Although only the low-energy conformations are generated and saved to $\L$, the number of all conformations can still be large for
ligands with many rotatable bonds.
The aim of this experiment was to verify the procedure to select a subset of ligand conformations from $\L$.

Two strategies to select a subset from the $\L$ conformations were tested: 
the strategy described in Sec.~\ref{sec::strat} (denoted `S') and the strategy that prefers smaller (packed) conformations (denoted `P').
The `P' method selects such conformations where the deviations of the atomic distance from the geometric center are the lowest, which leads 
to the selection of packed conformations.
The strategies were tested using the ligand DCP (Dichloropropanol), for which  $|\L|=1300$ conformations were created, and for the ligand
m040 (1,5-Dichloropentane) with $|\L|=7560$ conformations.

@@

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
    The number after~`$/$' denotes the number of atoms.
    The protein has 4650 atoms. 
    The rate is shown in [\%].
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
%\renewcommand{\arraystretch}{0.7}
\input graphSelNew.tex
}
\end{table}

\subsubsection{Traversability of Ligands}

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
    The protein has 4650 atoms.
    The rate is shown in [\%].
}
\small
\renewcommand{\tabcolsep}{2pt}
{
\scriptsize
%\renewcommand{\arraystretch}{0.7}
%\input graphNew.tex %with \rc (a1*) column
\input graphNewNoA3.tex %without \rc (a1*) column
}
\end{table}

@@

\begin{table}[t]
\caption{\label{fig::m040c}
\small
Examples of elongated/packed conformation for m040 (1,5-Dichloropentane).
The table shows the average number of conformations (out of 50) used in trajectories reaching the active site.
}
\centering
{\footnotesize
%\def\arraystretch{0.9}
\begin{tabular}{cc}
\rotatebox{0}{\hskip 5pt Long} \\ \includegraphics[width=0.1\textwidth]{fig/m040-conf1} \\
\rotatebox{0}{Packed} \\ \includegraphics[width=0.078\textwidth]{fig/m040-conf2}
\end{tabular}
}
{\scriptsize
\def\arraystretch{0.9}
\input usedc.tex
}
\end{table}

\subsection{Analysis of protein conformations}

The next studied case was the analysis of Leukotriene A4 hydrolase (PDB ID 4L2L).
The task was to find trajectory for the ligand 4-(4-benzylphenyl)-1,3-thiazol-2-amine, which is a known inhibitor of the protein 4L2L.
We therefore assume that the ligand can reach the active site.
%which was in the protein crystal structure (PDB ID 4L2L). 
%Based on the location of inhibitor inside the protein we assume that the inhibitor is able 
The tCONCOORD tool~\cite{seeliger2007geometry} was used to generate 1000 conformations of the protein 4L2L (without the inhibitor).
The initial pdb structure was minimized by GROMACS 4.0.7~\cite{hess2008gromacs} using the input parameters recommended on the example page for tCONCOORD. 
The protein conformations were generated using default settings for tCONCOORD in the PERT mode. % which only perturbs the starting configuration of atoms instead of complete randomization.
The tunnels towards the active site (defined by residues 134 (Alanine) and 311 (Phenylalanine) for probe 0.9~\AA\ were calculated
using CAVER 3.0.
All 1000 conformations contained at least 5 tunnels, more than 900 conformations contained 9 tunnels.
The ordering of the tunnels can slightly change due to conformational changes, so the traversability
to the active site was computed for the first three tunnels (with average bottlenecks 1.22~\AA, 1.20~\AA and 1.19~\AA).
Due to size of the ligand (which has 33 atoms) and the low bottlenecks of the tunnels, we

%R1: https://www.ncbi.nlm.nih.gov/pubmed/17997973
%R2: https://www.ncbi.nlm.nih.gov/pubmed/26620784

\begin{figure}
\centering
\caption{\label{fig::inhibitor}Example of inhibitor in 4L2L ... }
\end{figure}

@@

%originally last two paragraphs in the experiment section, for resubmition, they are moved here to discussion
The proposed \RA\ planner provides higher traversability rate than RRT-Retraction (\RB).
The distances of the configuration trees (measured using the 3D Euclidean metric) to the active site and the runtimes in each frame
are shown in Fig.~\ref{fig::comparison}.
The runtimes of \RB\ are higher than runtimes of \RA\ and the std. deviations as well.
%The variant \RC\ is faster than \RA\ due to simple evaluation of the metric in the expansion step.
%The runtimes of both algorithms increase in those frames, where the tunnel is not traversable, i.e., where the distance
%to the active site is more than 2~\AA\ (for example, in the frame no. 60).
%The utilization of the  atom-based metric (used in \RA) leads to higher traversability rate
%in the case of large scales and larger ligands (e.g., m056 with $\smin=0.8$: 15~\% with the metric, but to only 6~\% without it).
%Both \RA\ and \RC\ provided trajectories with the same traversability rate in the case of lower scales ($\smin<0.6$).

@@

We presented a novel modification of the RRT planner to compute trajectories of ligands inside selected protein tunnels.
In our solution, we take into account the ligand shape and we model the ligand flexibility using a predefined set of its conformations.
To simulate the real behavior of ligand passing through a tunnel, we enable scaling down the atomic radii of the ligands up to a certain predefined factor.
The planner generates the random samples only along a tunnel computed by the Voronoi diagram, and it attempts to expand the configuration
tree using all ligand conformations. 
The expansion prefers atoms with a higher scale, which ensures that the ligand shrinks down only if necessary, i.e., when traversing the narrow parts of the tunnel.
%To further boost the exploration of the configuration space, a novel metric suitable for measuring the distances between various ligand shapes is used in the expansion step.
%The proposed method has been used to test and evaluate the trajectories of several ligands in a sequence of frames capturing protein dynamics.
